---
title: "Deconvolution of gene expression Ewing Sarcoma datasets"
author: "Jane Merlevede & Andrei Zinovyev"
date: "`r format(Sys.time(), '%m/%d/%y')`"  
output:
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 4
    number_sections: true  
    df_print: paged 
---
  

```{r variable_definition, echo=FALSE}
#Options to be defined by the user
tumor_type="ES"
nPerm=10000 #Nb permutations for gsea
top_driver_genes=5 #Nb of top top-contributing genes to show in table
```

```{r file_loading, echo=FALSE} 
#input files to provide
weighted_metagenes = read.table(paste0(tumor_type,"_weighted_metagenes"), header = TRUE, sep="\t") #weighted metagenes computed in mcl_and_weighted_metagenes.R
datasets_info = read.table("../datasets_info", header = TRUE, sep="\t", as.is=TRUE) #info about the datasets
community_comp = read.table(paste0(tumor_type,"_communities"), header = TRUE, sep="\t") #obtained from mcl_and_weighted_metagenes.R
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package_loading, echo=FALSE, message = FALSE}
library("stringr")
library("knitr")
library("clusterProfiler")
library(GOSemSim)
library("enrichplot")
library("vroom")
library("DOSE")
library("ggplot2")
library("cowplot")
library("DT")
```


# Introduction

We described here the analyses performed on `r tumor_type` gene expression datasets. Our goal is to identify robust (across different datasets) mechanims active in `r tumor_type` tumors.

This report described in details the communities identified: their component names, their techno type (bulk or single), their data type (patient, PDX, cell line, controls) and the significantly enriched gene sets. It also briefly describes the main methodological steps of the approach. 

The inputs needed are:
* weighted metagenes previously computed
* the community composition obtained after the computation of weighted metagenes
* information about the datasets


This work is in the scope of D3.1 iPC deliverable.

Here we focused on Ewing sarcoma, for which we gathered 20 datasets: 

* 5 bulk datasets of patients
* 4 single cell cell lines
* 8 single cell PDX
* 3 single cell controls (including 2 full embryos GSE137804)

```{r dataset_infos, echo=FALSE}
datasets_info_ES = subset(datasets_info, tumor_type=="ES")
datasets_info_ctrl=subset(datasets_info, tumor_type=="Ctrl")
data_infos=rbind(datasets_info_ES, datasets_info_ctrl)
data_infos=data_infos[,-1]
kable(data_infos[,c(4,1:3)])
```

# Description of the communities

## Community composition
In `r tumor_type` datasets, `r nrow(community_comp)` communities containing at least 3 distinct datasets were identified. The number of components per community and the community composition are shown below.

```{r table_community_composition, echo=FALSE}
barplot(rowSums(!is.na(community_comp)), xlab="Community", ylab="Nb of components", main="Nb of components per community", names.arg = 1:nrow(community_comp))
datatable(community_comp)
```
Now, we want to characterize these communities by the nature of the components they contain and by the mechanims they represent. 

## Nature of the communities
First, for each community, we describe how many components they contain and from how many unique datatsets they come from. 
We also show how many bulk and single cell data were gathered. The sample type, patients, cell lines, PDX and controls is also given and is a first clue for the interpretation.

```{r datasets_info, echo=FALSE} 
community_stats=as.data.frame(matrix(0, nrow = nrow(community_comp), ncol = 0))
community_stats$community=paste0("C",1:nrow(community_comp))
community_stats$nb_comp = rowSums(!is.na(community_comp))

for (c in 1:nrow(community_comp))
{
  current_community = community_comp[c,!is.na(community_comp[c,])]
  comp_name=rep(0,length(current_community))
  for (k in 1:length(current_community))
  {
    comp_name[k] = unlist(str_split(current_community[1,k],"_metagene"))[1]
  }
  
  ind_comp_current_community = which(datasets_info$Acronym %in% comp_name)
  datasets_info_current_community = datasets_info[ind_comp_current_community,]
  
  community_stats$nb_comp_unique[c] = length(unique(comp_name))
  community_stats$nb_bulk[c]= sum(datasets_info_current_community$Techno=="bulk")
  community_stats$nb_sc[c]= sum(datasets_info_current_community$Techno=="single-cell")
  community_stats$nb_patient[c] = sum(datasets_info_current_community$data_type=="Patient")
  community_stats$nb_cellline[c] = sum(datasets_info_current_community$data_type=="cell-line")
  community_stats$nb_PDX[c] = sum(datasets_info_current_community$data_type=="PDX")
  community_stats$nb_control[c] = sum(datasets_info_current_community$data_type=="control")  
}
kable(community_stats)
```

```{r datasets_info_to_mention_in_text, echo=FALSE, message=FALSE, warning=FALSE} 
nb_bulk_only=length(which(community_stats$nb_sc == 0))
nb_sc_only=length(which(community_stats$nb_bulk == 0))
nb_no_control=length(which(community_stats$nb_control == 0))
```

`r nrow(community_stats)-nb_bulk_only-nb_sc_only` out of `r nrow(community_stats)` communities are a mixture of bulk and single cell data, `r nb_bulk_only` contains only bulk and `r nb_sc_only` contains only single cell datasets.

`r nb_no_control` out of `r nrow(community_stats)` communities does not contain any control. As we want to focuse on tumor related signals, we can exclude communities containing at least one component from control datasets. Thus we further focuse on `r nb_no_control` communities, which are communities `r which(community_stats$nb_control == 0)`. !!! To discuss with Andrei !!!

## Top-contributing genes

The top-contributing genes (genes with the hightest and smallest weights) are shown below for each component.
```{r table_driver_genes, echo=FALSE, message=FALSE, warning=FALSE}
#for each weighted metagenes of interest, top-contributing genes are the genes beyond 3 SD from the mean
Mean=colMeans(weighted_metagenes, na.rm = TRUE)
SD=apply(weighted_metagenes,2,sd, na.rm = TRUE)
  
nb_pos_tail_3SD=nb_neg_tail_3SD=rep(0,ncol(weighted_metagenes))
scores_pos_tail_3SD=scores_neg_tail_3SD=rep(0,ncol(weighted_metagenes))
fn = paste0(tumor_type,"_top_contributing_genes")
if (file.exists(fn,showWarnings = FALSE)) {
  #Delete file if it exists
  file.remove(fn)
}
  
for (k in 1:ncol(weighted_metagenes))
{
  ind_pos_tail_3SD=which(weighted_metagenes[,k]>Mean[k]+3*SD[k])
  ind_neg_tail_3SD=which(weighted_metagenes[,k]<Mean[k]-3*SD[k])
  genes_pos_tail= rownames(weighted_metagenes[ind_pos_tail_3SD,])
  genes_neg_tail= rownames(weighted_metagenes[ind_neg_tail_3SD,])
  genes=c(genes_pos_tail, genes_neg_tail)
  scores=c(weighted_metagenes[ind_pos_tail_3SD,k], weighted_metagenes[ind_neg_tail_3SD,k])
  driver_genes=data.frame(cbind(paste0("C",k), genes, scores))
  write.table(driver_genes , paste0(tumor_type,"_top-contributing_genes"), sep="\t", row.names = FALSE, append=TRUE, 
              col.names=!file.exists( paste0(tumor_type,"_top-contributing_genes")), quote=FALSE)
}

driver_genes = read.table(paste0(tumor_type,"_top-contributing_genes"), header = TRUE, sep="\t", as.is=TRUE)
table_top_drivers=data.frame(Community=numeric(nrow(community_comp)), Driver_pos_tail=numeric(nrow(community_comp)), Driver_neg_tail=numeric(nrow(community_comp)))
for (i in 1:nrow(community_comp))
{
  community = subset(driver_genes, V1==paste0("C",i))
  community_pos = community[which(community$scores>=0),]
  community_neg = community[which(community$scores<0),]  
  community_pos = community_pos[order(community_pos$scores,decreasing=TRUE),]
  community_neg = community_pos[order(community_neg$scores,decreasing=FALSE),]
  
  table_top_drivers$Community[i] <- paste0("C",i)
  table_top_drivers$Driver_pos_tail[i] = paste0(community_pos$genes[1:top_driver_genes], collapse=";")
  table_top_drivers$Driver_neg_tail[i] = paste0(community_neg$genes[1:top_driver_genes], collapse=";")
}
datatable(table_top_drivers, rownames=FALSE)
```

# Annotation of the communities using GO, KEGG, MSigDB and DO

To decipher the mechnanisms active in the communities, we extracted the top-contributing genes in each community and performed over representation analysis as well as gene set enrichment analyses using MSigDB (http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp) on each community.

## Over representation analysis of the top-contributing genes and gene set enrichment analysis for each community

Now we can describe each component in details. We used GO, KEGG and DO ontology to annotate the communities. Results are represented in dotplots, cnetplots ans emaplots for both drviers of the positive and negative tails using clusterProfiler (https://bookdown.org/yihui/rmarkdown/html-document.html). If the OR analysis had no enrichment in one or the two tails, no plot are reported. We used GO, KEGG and disease ontology databases.

```{r communities_OR_details, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
gcSample_pos = list()
gcSample_neg = list()
Mean=colMeans(weighted_metagenes, na.rm = TRUE)
SD=apply(weighted_metagenes,2,sd, na.rm = TRUE)

folder_MSigDB_gene_sets = "gene_sets/MSigDB/"
gene_sets = list.files(folder_MSigDB_gene_sets, full.names = TRUE)
egmt_pos_combine=data.frame(matrix(nrow=0,ncol=10))
egmt_neg_combine=data.frame(matrix(nrow=0,ncol=10))
d <- godata('org.Hs.eg.db', ont="BP")
  
out <- NULL

for (i in 1:nrow(community_comp)) #nrow(community_comp)
{
  out <- c(out, knit_child('template_pathway_analysis.Rmd', quiet = TRUE))
} 
```

`r paste(out, collapse='\n')`


## Overview of over representation analysis in top-contributing genes of all communities
top-contributing genes were submitted to over representation analysis. First, we show an overview of the top-contributing genes coming from the postive and negative tails for all components and next, we give details about each component.

```{r communities_overview, echo=FALSE, message = FALSE, warning = FALSE, fig.height=30, fig.width=40}
#, fig.cap=c("KEGG overrepresentation of positive tail", "KEGG overrepresentation of negative tail", "GO overrepresentation of positive tail", "GO overrepresentation of negative tail", "DO overrepresentation of positive tail", "DO overrepresentation of negative tail")
#,fig.fullwidth = TRUE
gcSample_pos = list()
gcSample_neg = list()
Mean=colMeans(weighted_metagenes, na.rm = TRUE)
SD=apply(weighted_metagenes,2,sd, na.rm = TRUE)
for (i in 1:nrow(community_comp)) #nrow(community_comp)
{
 ## feature 1: numeric vector
 geneList <- weighted_metagenes[,i]
 ## feature 2: named vector
 names(geneList) <- rownames(weighted_metagenes)

 ## feature 3: decreasing order
 geneList <- sort(geneList, decreasing = TRUE)
 #data(geneList, package="DOSE")

 #define top-contributing genes
 ind_pos_tail_3SD=which(weighted_metagenes[,i]>Mean[i]+3*SD[i])
 ind_neg_tail_3SD=which(weighted_metagenes[,i]<Mean[i]-3*SD[i])
 genes_pos_tail= rownames(weighted_metagenes[ind_pos_tail_3SD,])
 genes_neg_tail= rownames(weighted_metagenes[ind_neg_tail_3SD,])
 genes=c(genes_pos_tail, genes_neg_tail)

 gene_pos_tail = weighted_metagenes[ind_pos_tail_3SD,i]
 gene_neg_tail = weighted_metagenes[ind_neg_tail_3SD,i]
 names(gene_pos_tail) = rownames(weighted_metagenes[ind_pos_tail_3SD,])
 names(gene_neg_tail) = rownames(weighted_metagenes[ind_neg_tail_3SD,])
 gene_pos_tail_temp = sort(gene_pos_tail, decreasing = TRUE)
 gene_neg_tail_temp = sort(gene_neg_tail, decreasing = TRUE)
 gene_pos_tail_list = bitr(names(gene_pos_tail_temp), fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
 gene_neg_tail_list = bitr(names(gene_neg_tail_temp), fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")

 key_pos = paste0("C",i,"_p")
 key_neg = paste0("C",i,"_n")
 gcSample_pos[[ key_pos ]] <- gene_pos_tail_list$ENTREZID
 gcSample_neg[[ key_pos ]] <- gene_neg_tail_list$ENTREZID
}

ck_pos <- compareCluster(geneCluster = gcSample_pos, fun = "enrichKEGG")
ck_neg <- compareCluster(geneCluster = gcSample_neg, fun = "enrichKEGG")
cGO_pos <- compareCluster(geneCluster = gcSample_pos, fun = "enrichGO", OrgDb = org.Hs.eg.db)
cDO_pos <- compareCluster(geneCluster = gcSample_pos, fun = "enrichDO")
cGO_neg <- compareCluster(geneCluster = gcSample_neg, fun = "enrichGO", OrgDb = org.Hs.eg.db)
cDO_neg <- compareCluster(geneCluster = gcSample_neg, fun = "enrichDO")

dotplot(cGO_pos,font.size=32, title="GO overrepresentation for the top-contributing genes of the positive tail")
cat('\n\n')
dotplot(cGO_neg,font.size=32, title="GO overrepresentation for the top-contributing genes of the negative tail")
cat('\n\n')
dotplot(ck_pos,font.size=32, title="KEGG overrepresentation for the top-contributing genes of the positive tail")
cat('\n\n')
dotplot(ck_neg,font.size=32, title="KEGG overrepresentation for the top-contributing genes of the negative tail")
cat('\n\n')
dotplot(cDO_pos,font.size=32, title="DO overrepresentation for the top-contributing genes of the positive tail")
cat('\n\n')
dotplot(cDO_neg,font.size=32, title="DO overrepresentation for the top-contributing genes of the negative tail")
```


# Annotation of the communities using gene sets of transcription factors
We used Dorothea database (Discriminant Regulon Expression Analysis, https://saezlab.github.io/dorothea/, Garcia-Alonso L et al. 2019) to construct gene sets of each of the reported transcription factor. We then checked if the communities were enriched in some of these TF gene sets.

For that, we performed over representation analysis of the top-contributing genes of the positive and negative tails of each community using the TF gene sets.


```{r communities_OR_TF_gene_sets, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}

TF = "gene_sets/TF_targets_dorothea_A-D_hs.gmt"
TFgs <- read.gmt(TF)
egmt_pos_combine=data.frame(matrix(nrow=0,ncol=10))
egmt_neg_combine=data.frame(matrix(nrow=0,ncol=10))

out_TFgs <- NULL

for (i in 1:nrow(community_comp) ) 
{
  out_TFgs <- c(out_TFgs, knit_child('template_enrichment_in_TF_genesets.Rmd', quiet = TRUE))
}
```

`r paste(out_TFgs, collapse='\n')`

```{r communities_OR_TF_gene_sets_datatable, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
datatable(egmt_pos_combine_reordered, rownames = FALSE)
datatable(egmt_neg_combine_reordered, rownames = FALSE)
```

For each community, a total of `r length(unique(TFgs$term))` gene sets are tested. 

There are a total of `r nrow(egmt_pos_combine_reordered)` significant enrichments in positive tails and `r nrow(egmt_neg_combine_reordered)` significant enrichment in negative tails.

There are `r length(unique(egmt_pos_combine_reordered$ID))`  unique TFs with significant enrichments in their targets in the top-contributing genes of the positive tails and `r length(unique(egmt_neg_combine_reordered$ID))` unique TFs with significant enrichments in their targers in the top-contributing genes of the negative tails.



# Annotation of the communities using user defined gene sets

Now we want to check if the detected communities are enriched in specific gene sets. Here we used the gene sets identified in the work of Aynaud et al. 2019, where single cells ES data were investiguated using ICA. Also, additional gene sets are used.

For that, we performed over representation analysis of the top-contributing genes of the positive and negative tails of each community using the defined gene sets.

```{r communities_OR_personnalized_gene_sets, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}

specific_gene_sets = "gene_sets/ics_and_signatures.gmt"
sgs <- read.gmt(specific_gene_sets)
egmt_pos_combine=data.frame(matrix(nrow=0,ncol=10))
egmt_neg_combine=data.frame(matrix(nrow=0,ncol=10))

out_sgs <- NULL

for (i in 1:nrow(community_comp) ) 
{
  out_sgs <- c(out_sgs, knit_child('template_enrichment_in_specific_genesets.Rmd', quiet = TRUE))
}
```


`r paste(out_sgs, collapse='\n')`

```{r communities_OR_personnalized_gene_sets_datatable, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
datatable(egmt_pos_combine_reordered, rownames = FALSE)
datatable(egmt_neg_combine_reordered, rownames = FALSE)

egmt_pos_combine_reordered_ID=egmt_pos_combine_reordered[grep("^IC",egmt_pos_combine_reordered$ID),]
egmt_neg_combine_reordered_ID=egmt_neg_combine_reordered[grep("^IC",egmt_neg_combine_reordered$ID),]

unique_IC_pos = length( union( unique(egmt_pos_combine_reordered_ID$ID[grep("[+]",egmt_pos_combine_reordered_ID$ID)]), unique(egmt_neg_combine_reordered_ID$ID[grep("[+]",egmt_neg_combine_reordered_ID$ID)])) )

unique_IC_neg = length( union( unique(egmt_pos_combine_reordered_ID$ID[grep("-",egmt_pos_combine_reordered_ID$ID)]), unique(egmt_neg_combine_reordered_ID$ID[grep("-",egmt_neg_combine_reordered_ID$ID)]) ) )
```

For each community, a total of `r length(unique(sgs$term))` gene sets are tested. 

There are a total of `r nrow(egmt_pos_combine_reordered)` significant enrichments in positive tails and `r nrow(egmt_neg_combine_reordered)` significant enrichment in negative tails.

`r length(grep("^IC",egmt_pos_combine_reordered$ID))` IC are retrieved in positive tails and `r length(grep("^IC",egmt_neg_combine_reordered$ID))` IC are retrieved in negative tails.

`r length(unique(egmt_pos_combine_reordered$ID[grep("^IC",egmt_pos_combine_reordered$ID)]))` IC (`r length(unique(egmt_pos_combine_reordered_ID$ID[grep("[+]",egmt_pos_combine_reordered_ID$ID)]))` IC+ and `r length(unique(egmt_pos_combine_reordered_ID$ID[grep("-",egmt_pos_combine_reordered_ID$ID)]))` IC-) are retrieved in positive tails and `r length(unique(egmt_neg_combine_reordered$ID[grep("^IC",egmt_neg_combine_reordered$ID)]))` IC (`r length(unique(egmt_neg_combine_reordered_ID$ID[grep("[+]",egmt_neg_combine_reordered_ID$ID)]))` IC+ and `r length(unique(egmt_neg_combine_reordered_ID$ID[grep("-",egmt_neg_combine_reordered_ID$ID)]))` IC-) are retrieved in negative tails.

`r unique_IC_pos` IC+ and `r unique_IC_neg` IC- are retrived.