

```{r, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
## feature 1: numeric vector
geneList <- weighted_metagenes[,i]
## feature 2: named vector
names(geneList) <- rownames(weighted_metagenes)

## feature 3: decreasing order
geneList <- sort(geneList, decreasing = TRUE)
geneList_HUGO = geneList
#data(geneList, package="DOSE")
geneList_conversion = bitr(names(geneList), fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
names(geneList) <- geneList_conversion$ENTREZID
names(geneList_HUGO) <- geneList_conversion$SYMBOL


#define driver genes
ind_pos_tail_3SD=which(weighted_metagenes[,i]>Mean[i]+3*SD[i])
ind_neg_tail_3SD=which(weighted_metagenes[,i]<Mean[i]-3*SD[i])
gene_pos_tail = weighted_metagenes[ind_pos_tail_3SD,i]
gene_neg_tail = weighted_metagenes[ind_neg_tail_3SD,i]
names(gene_pos_tail) = rownames(weighted_metagenes[ind_pos_tail_3SD,])
names(gene_neg_tail) = rownames(weighted_metagenes[ind_neg_tail_3SD,])
genes=c(gene_pos_tail, gene_neg_tail)
gene_pos_tail_temp = sort(gene_pos_tail, decreasing = TRUE)
gene_neg_tail_temp = sort(gene_neg_tail, decreasing = TRUE)
gene_pos_tail_list = bitr(names(gene_pos_tail_temp), fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
gene_neg_tail_list = bitr(names(gene_neg_tail_temp), fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")

#for the overview
 key_pos = paste0("C",i,"_p")
 key_neg = paste0("C",i,"_n")
 gcSample_pos[[ key_pos ]] <- gene_pos_tail_list$ENTREZID
 gcSample_neg[[ key_pos ]] <- gene_neg_tail_list$ENTREZID
```

### `r paste0("Community ",i)`

#### `r paste0("GO enrichment for community ",i)`
Let us take a look at the results of the enrichments of the drivers of the positive and negative tails based on Gene Ontology.

```{r, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
#GO
  ego_pos <- enrichGO(gene      = gene_pos_tail_list$ENTREZID,
                    universe      = names(geneList),
                    OrgDb         = org.Hs.eg.db,
                    #keyType = "SYMBOL",
                    ont           = "ALL",
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.01,
                    qvalueCutoff  = 0.01,
                    readable      = TRUE,
                    pool         = TRUE)

ego_neg <- enrichGO(gene      = gene_neg_tail_list$ENTREZID,
                    universe      = names(geneList),
                    OrgDb         = org.Hs.eg.db,
                    #keyType = "SYMBOL",
                    ont           = "ALL",
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.01,
                    qvalueCutoff  = 0.01,
                    readable      = TRUE)

datatable(data.frame(ego_pos), rownames=FALSE, caption=paste0("\n\n\nGO overexpression analysis for drivers (pos tail) in community ",i) ) 
datatable(data.frame(ego_neg), rownames=FALSE, caption=paste0("\n\n\nGO overexpression analysis for drivers (neg tail) in community ",i)) 

if (nrow(ego_pos) > 0 & nrow(ego_neg) == 0 )
{
  go1_d <- dotplot(ego_pos, showCategory=20) + ggtitle(paste0("\n\n\ndotplot of the drivers (pos tail) in community ",i))
  print(plot_grid(go1_d, ncol=1)) 
  if (nrow(ego_pos) > 1)
  {
    ego_pos2 <- pairwise_termsim(ego_pos, method="Wang", semData = d)
    go1_e <- emapplot(ego_pos2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot of the drivers (pos tail) in community ",i))
    print(plot_grid(go1_e, ncol=1)) #, labels=LETTERS[1:4]
  }
  go1_c <- cnetplot(ego_pos, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot of the drivers (pos tail) in community ",i))
  print(plot_grid(go1_c, ncol=1)) #, labels=LETTERS[1:4]
} else if(nrow(ego_pos) == 0 & nrow(ego_neg) > 0 )
{
  go2_d <- dotplot(ego_neg, showCategory=20) + ggtitle(paste0("\n\n\ndotplot of the drivers drivers (neg tail) in community ",i))
  print(plot_grid(go2_d, ncol=1)) 
  if (nrow(ego_neg) > 1)
  {
    ego_neg2 <- pairwise_termsim(ego_neg, method="Wang", semData = d)
    go2_e <- emapplot(ego_neg2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot of the drivers (neg tail) in community ",i))
    print(plot_grid(go2_e, ncol=1)) #, labels=LETTERS[1:4]  
  }
  go2_c <- cnetplot(ego_neg, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot of the drivers (neg tail) in community ",i))
  print(plot_grid(go2_c, ncol=1)) #, labels=LETTERS[1:4]
} else if (nrow(ego_pos) > 0 & nrow(ego_neg) > 0 )
{
  go1_d <- dotplot(ego_pos, showCategory=20, x = "GeneRatio") + ggtitle(paste0("\n\n\ndotplot of the drivers drivers (pos tail) in community ",i))
  go2_d <- dotplot(ego_neg, showCategory=20, x = "GeneRatio") + ggtitle(paste0("\n\n\ndotplot of the drivers drivers (neg tail) in community ",i))
  print(plot_grid(go1_d,go2_d, ncol=2)) 
  if (nrow(ego_pos) > 1 & nrow(ego_neg) > 1 )
  {
    ego_pos2 <- pairwise_termsim(ego_pos, method="Wang", semData = d)
    ego_neg2 <- pairwise_termsim(ego_neg, method="Wang", semData = d)
    go1_e <- emapplot(ego_pos2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot of the  drivers (pos tail) in community ",i))
    go2_e <- emapplot(ego_neg2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot of the  drivers (neg tail) in community ",i))
    print(plot_grid(go1_e, go2_e, ncol=2)) #, labels=LETTERS[1:4]
  }
  go1_c <- cnetplot(ego_pos, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot of the drivers (pos tail) in community ",i))
  go2_c <- cnetplot(ego_neg, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot of the drivers (neg tail) in community ",i))
  print(plot_grid(go1_c, go2_c, ncol=2)) #, labels=LETTERS[1:4]
}


ego_gse <- gseGO(geneList     = geneList,
              OrgDb        = org.Hs.eg.db,
              #keyType = "SYMBOL",
              ont          = "ALL",
              nPerm        = nPerm, 
              minGSSize    = 10,
              maxGSSize    = 900,
              pvalueCutoff = 0.05,
              pAdjustMethod = "BH",
              verbose      = FALSE)

ego_gse <- setReadable(ego_gse, 'org.Hs.eg.db',keyType = "auto")
ind_NES=which(abs(ego_gse$NES)>=2)
ego_gse=ego_gse[ind_NES,]
ego_gse_pos=ego_gse[which(ego_gse$NES>=2),]
ind_order_pos=order(ego_gse_pos$NES, decreasing=TRUE)
ego_gse_pos=ego_gse_pos[ind_order_pos,]
ego_gse_neg=ego_gse[which(ego_gse$NES<=-2),]
ind_order_neg=order(ego_gse_neg$NES, decreasing=FALSE)
ego_gse_neg=ego_gse_neg[ind_order_neg,]

datatable(data.frame(ego_gse_pos), rownames=FALSE, caption=paste0("\n\n\nGO gsea for drivers (pos tail) in community ",i))
datatable(data.frame(ego_gse_neg), rownames=FALSE, caption=paste0("\n\n\nGO gsea for drivers (neg tail) in community ",i))

#go_gsea <- dotplot(ego_gse, showCategory=20) + ggtitle(paste0("GO - drivers (pos tail) in community ",i))
#print(plot_grid(go_gsea, ncol=1)) 
```

<br><br>

#### `r paste0("KEGG enrichment for community ",i)`
Let us now take a look at the results of the enrichments of the drivers of the positive and negative tails based on KEGG.

```{r, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
#KEGG
ek_pos <- enrichKEGG(gene = gene_pos_tail_list$ENTREZID,
                     organism     = 'hsa',
                     pvalueCutoff = 0.01)

ek_neg <- enrichKEGG(gene = gene_neg_tail_list$ENTREZID,
                     organism     = 'hsa',
                     pvalueCutoff = 0.01)

ek_pos <- setReadable(ek_pos, 'org.Hs.eg.db','ENTREZID')
ek_neg <- setReadable(ek_neg, 'org.Hs.eg.db', 'ENTREZID')

DT::datatable(data.frame(ek_pos), rownames=FALSE, caption=paste0("\n\n\nKEGG overexpression analysis for drivers (pos tail) in community ",i))
DT::datatable(data.frame(ek_neg), rownames=FALSE, caption=paste0("\n\n\nKEGG overexpression analysis for drivers (neg tail) in community ",i))

ek1_d <- dotplot(ek_pos, showCategory=20) + ggtitle(paste0("\n\n\ndotplot for the drivers (pos tail) in community ",i))
ek2_d <- dotplot(ek_neg, showCategory=20) + ggtitle(paste0("\n\n\ndotplot for the drivers (neg tail) in community ",i))

if (nrow(ek_pos) > 0 & nrow(ek_neg) == 0 )
{
  print(plot_grid(ek1_d, ncol=1))
  # if (nrow(ek_pos) > 1)
  # {
  #   ek1_e <- emapplot(ek_pos, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (pos tail) in community ",i))
  #   print(plot_grid(ek1_e, ncol=1)) #, labels=LETTERS[1:4]
  # }
  ek1_c <- cnetplot(ek_pos, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (pos tail) in community ",i))
  print(plot_grid(ek1_c, ncol=1)) #, labels=LETTERS[1:4]
} else if(nrow(ek_pos) == 0 & nrow(ek_neg) > 0 )
{
  print(plot_grid(ek2_d, ncol=1))
  # if (nrow(ek_neg) > 1)
  # {
  #   ek2_e <- emapplot(ek_neg, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (neg tail) in community ",i))
  #   print(plot_grid(ek2_e, ncol=1)) #, labels=LETTERS[1:4]
  # }
  ek2_c <- cnetplot(ek_neg, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (neg tail) in community ",i))
  print(plot_grid(ek2_c, ncol=1)) #, labels=LETTERS[1:4]
} else if (nrow(ek_pos) > 0 & nrow(ek_neg) > 0 )
{
  print(plot_grid(ek1_d,ek2_d, ncol=2))
   # if (nrow(ek_pos) > 1 & nrow(ek_neg) > 1 )
   # {
   #    ek1_e <- emapplot(ek_pos, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (pos tail) in community ",i))
   #    ek2_e <- emapplot(ek_neg, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (neg tail) in community ",i))
   #    print(plot_grid(ek1_e, ek2_e, ncol=2)) #, labels=LETTERS[1:4]
   # }
  ek1_c <- cnetplot(ek_pos, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (pos tail) in community ",i))
  ek2_c <- cnetplot(ek_neg, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (neg tail) in community ",i))
  print(plot_grid(ek1_c, ek2_c, ncol=2)) #, labels=LETTERS[1:4]
}


ek_gse <- gseKEGG(geneList     = geneList,
               organism     = 'hsa',
               nPerm        = nPerm,
               minGSSize    = 10,
               maxGSSize = 900,
               pvalueCutoff = 0.05,
               verbose      = FALSE)

ek_gse <- setReadable(ek_gse, 'org.Hs.eg.db',keyType = "ENTREZID")
ind_NES=which(abs(ek_gse$NES)>=2)
ek_gse=ek_gse[ind_NES,]
ek_gse_pos=ek_gse[which(ek_gse$NES>=2),]
ind_order_pos=order(ek_gse_pos$NES, decreasing=TRUE)
ek_gse_pos=ek_gse_pos[ind_order_pos,]
ek_gse_neg=ek_gse[which(ek_gse$NES<=-2),]
ind_order_neg=order(ek_gse_neg$NES, decreasing=FALSE)
ek_gse_neg=ek_gse_neg[ind_order_neg,]

datatable(data.frame(ek_gse_pos), rownames=FALSE, caption=paste0("\n\n\nKEGG gsea for drivers (pos tail) in community ",i))
datatable(data.frame(ek_gse_neg), rownames=FALSE, caption=paste0("\n\n\nKEGG gsea for drivers (neg tail) in community ",i))

```


<br><br>

#### `r paste0("MSigDB enrichment for community ",i)`
Let us now take a look at the results of the enrichments of the drivers of the positive and negative tails based on MSigDB catalogues.


```{r, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
egmt_pos_combine=data.frame(matrix(nrow=0,ncol=10))
egmt_neg_combine=data.frame(matrix(nrow=0,ncol=10))

for (gmt_files in gene_sets)
{
  c <- read.gmt(gmt_files)
  msigdb_egmt_pos <- enricher(gene_pos_tail_list$SYMBOL, TERM2GENE=c, pvalueCutoff = 0.01, pAdjustMethod = "BH", universe=names(geneList_HUGO), minGSSize = 10, maxGSSize = 900, qvalueCutoff = 0.01)
  msigdb_egmt_neg <- enricher(gene_neg_tail_list$SYMBOL, TERM2GENE=c, pvalueCutoff = 0.01, pAdjustMethod = "BH", universe=names(geneList_HUGO), minGSSize = 10, maxGSSize = 900, qvalueCutoff = 0.01)
  
  if ( nrow(msigdb_egmt_pos)>0 )
  {
    egmt_pos = as.data.frame(msigdb_egmt_pos)
    egmt_pos$Community=paste0("C",i)
    egmt_pos_combine=rbind(egmt_pos_combine,egmt_pos)
  }
  
  if ( nrow(msigdb_egmt_neg)>0 )
  {
    egmt_neg = as.data.frame(msigdb_egmt_neg)
    egmt_neg$Community=paste0("C",i)
    egmt_neg_combine=rbind(egmt_neg_combine,egmt_neg)
  }
}

datatable(egmt_pos_combine[,c(10,1:9)], rownames = FALSE)
datatable(egmt_neg_combine[,c(10,1:9)], rownames = FALSE)
```


<br><br>

#### `r paste0("DO enrichment for community ",i)`
Finally, let us take a look at the results of the enrichments of the drivers of the positive and negative tails based on Disease Ontology.

```{r, echo=FALSE, message=FALSE, warning = FALSE,fig.fullwidth = TRUE, fig.width=12}
#DO
edo_pos <- enrichDO(gene  = gene_pos_tail_list$ENTREZID,
                    ont           = "DO",
                    pvalueCutoff  = 0.01,
                    pAdjustMethod = "BH",
                    universe      = names(geneList),
                    minGSSize     = 5,
                    maxGSSize     = 500,
                    qvalueCutoff  = 0.01,
                    readable      = FALSE)

edo_neg <- enrichDO(gene  = gene_neg_tail_list$ENTREZID,
                    ont           = "DO",
                    pvalueCutoff  = 0.01,
                    pAdjustMethod = "BH",
                    universe      = names(geneList),
                    minGSSize     = 5,
                    maxGSSize     = 500,
                    qvalueCutoff  = 0.01,
                    readable      = FALSE)

edo_pos <- setReadable(edo_pos, 'org.Hs.eg.db')
edo_neg <- setReadable(edo_neg, 'org.Hs.eg.db')


DT::datatable(data.frame(edo_pos), rownames=FALSE, caption=paste0("\n\n\nDO overexpression analysis for drivers (pos tail) in community ",i))
DT::datatable(data.frame(edo_neg), rownames=FALSE, caption=paste0("\n\n\nDO overexpression analysis for drivers (neg tail) in community ",i))

if (nrow(edo_pos) > 0 & nrow(edo_neg) == 0 )
{
  do1_d <- dotplot(edo_pos, showCategory=20) + ggtitle(paste0("\n\n\ndotplot for the drivers (pos tail) in community ",i))
  print(plot_grid(do1_d, ncol=1))
  # if (nrow(edo_pos) > 1)
  # {
  #   edo_pos2 <- pairwise_termsim(edo_pos, method="Wang", semData = do)
  #   do1_e <- emapplot(edo_pos2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (pos tail) in community ",i))
  #   print(plot_grid(do1_e, ncol=1)) #, labels=LETTERS[1:4]    
  # }
  do1_c <- cnetplot(edo_pos, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the  drivers (pos tail) in community ",i))
  print(plot_grid(do1_c, ncol=1)) #, labels=LETTERS[1:4]
} else if(nrow(edo_pos) == 0 & nrow(edo_neg) > 0 )
{
  do2_d <- dotplot(edo_neg, showCategory=20) + ggtitle(paste0("\n\n\ndotplot for the drivers (neg tail) in community ",i))
  print(plot_grid(do2_d, ncol=1))
  # if (nrow(edo_neg) > 1)
  # {
  #   edo_neg2 <- pairwise_termsim(edo_neg, method="Wang", semData = do)
  #   do2_e <- emapplot(edo_neg2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (neg tail) in community ",i))
  #   print(plot_grid(do2_e, ncol=1)) #, labels=LETTERS[1:4]
  # }
  do2_c <- cnetplot(edo_neg, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (neg tail) in community ",i))
  print(plot_grid(do2_c, ncol=1)) #, labels=LETTERS[1:4]
} else if (nrow(edo_pos) > 0 & nrow(edo_neg) > 0 )
{
  do1_d <- dotplot(edo_pos, showCategory=20) + ggtitle(paste0("\n\n\ndotplot for the drivers (pos tail) in community ",i))
  do2_d <- dotplot(edo_neg, showCategory=20) + ggtitle(paste0("\n\n\ndotplot for the drivers (neg tail) in community ",i))
  print(plot_grid(do1_d,do2_d, ncol=2))
  # if (nrow(edo_pos) > 1 & nrow(edo_neg) > 1 )
  # {
  #   edo_pos2 <- pairwise_termsim(edo_pos, method="Wang", semData = do)
  #   edo_neg2 <- pairwise_termsim(edo_neg, method="Wang", semData = do)
  #   do1_e <- emapplot(edo_pos2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (pos tail) in community ",i))
  #   do2_e <- emapplot(edo_neg2, showCategory = 10) + ggtitle(paste0("\n\n\nemapplot for the drivers (neg tail) in community ",i))
  #   print(plot_grid(do1_e, do2_e, ncol=2)) #, labels=LETTERS[1:4]
  # }
  do1_c <- cnetplot(edo_pos, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (pos tail) in community ",i))
  do2_c <- cnetplot(edo_neg, foldChange=geneList, node_label="all") + ggtitle(paste0("\n\n\ncnetplot for the drivers (neg tail) in community ",i))
  print(plot_grid(do1_c, do2_c, ncol=2)) #, labels=LETTERS[1:4]
}


edo_gse <- gseDO(geneList,
           nPerm         = nPerm,
           minGSSize     = 10,
           maxGSSize = 900,
           pvalueCutoff  = 0.05,
           pAdjustMethod = "BH",
           verbose       = FALSE)

edo_gse <- setReadable(edo_gse, 'org.Hs.eg.db',keyType = "auto")
ind_NES=which(abs(edo_gse$NES)>=2)
edo_gse=edo_gse[ind_NES,]
edo_gse_pos=edo_gse[which(edo_gse$NES>=2),]
ind_order_pos=order(edo_gse_pos$NES, decreasing=TRUE)
edo_gse_pos=edo_gse_pos[ind_order_pos,]
edo_gse_neg=edo_gse[which(edo_gse$NES<=-2),]
ind_order_neg=order(edo_gse_neg$NES, decreasing=FALSE)
edo_gse_neg=edo_gse_neg[ind_order_neg,]

datatable(data.frame(edo_gse_pos), rownames=FALSE, caption=paste0("\n\n\nDO gsea for drivers (pos tail) in community ",i))
datatable(data.frame(edo_gse_neg), rownames=FALSE, caption=paste0("\n\n\nDO gsea for drivers (neg tail) in community ",i))
```


<br><br><br>